Individual and Differential Harm in Redistricting

### 2022-08-30 paper/harm.Rmd
draft edits

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank Kosuke Imai, Gary King, Shiro Kuriwaki, 
      Tyler Simko, Nick Stephanopolous, and 
      participants of the Fall 2021 Harvard American Politics Research Workshop 
      and Imai Research Group for helpful comments and discussion.
      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: apsr
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
First, it operates at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set
of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

@@

A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022]. The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}. Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

@@

```{r minissouri, fig.width=6.75}
#| fig.cap: |
#|      Two potential districting plans for Minissouri,
#|      a Democratic gerrymander (a), and a close-to-proportional plan (b).
#|      Panel (c) shows which voters are harmed by the Democratic gerrymander,
#|      using the proportional plan (b) as a counterfactual.
#|      Panel (d) summarizes this across precincts, showing
#|      the average harm done to voters of each partisan and racial group.
include_graphics("figures/minissouri.pdf")
```

@@

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent 44% of all Republicans and 2.5% of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of \$-20\\%\$, where positive values indicate pro-square bias.
In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b), and it does so on the same scale as differential partisan harm.

@@

A voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.

Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
In general, $q$ harms some individuals while $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals.
A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.

@@

Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$. Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
For algorithmic fairness, the implicit counterfactual is a perfect predictor---compared to this perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
Appendix \ref{app:elecmodel} contains more details about how we estimate and apply this model.

More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
For example, one could operate on the logit scale, or add terms to model incumbency effects.

@@

With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.

@@

A primary advantage of our proposed harm measures is that they our measures are defined at the individual level, which allows for the study of redistricting far beyond racial and partisan gerrymandering.
However, even for these important contexts, the measures provide an improvement over existing analytical approaches.

@@

The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
As such, no simple formulas exist to measure the precise racial impacts of a districting plan.
Indeed, much of the academic literature has focused on the statistical challenge of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].

The modern analysis of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.

While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, it is notable that the proposed harm measures unify all three *Gingles* criteria.
Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
But the proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.

@@

In both these cases, the proposed harm measures will quantify the extent to which a challenged districting plan differentially harms minority voters, taking into account their geographic and political cohesion, and the individual preferences of every voter.
Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.

@@

[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.

@@

Since none of the existing measures are centered on individual voters, they offer far less granular insight than the proposed harm framework, even in the case of studying partisan gerrymandering.
For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.

@@

Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
In the worst cases, this can lead to directionally wrong conclusions, as we will see in the next section.

@@

Figure \ref{fig:nj-pairs} in the appendix expands on Figure \ref{fig:nj-meas} to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.

@@

To apply our proposed framework to study the council district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
To estimate election outcomes, we require some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and thus must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu opponent Annissa Essaibi-George on the right.
The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.

Like all ecological inference models, ours rests on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not strictly met.
Yet we believe that the model is a useful simplification of reality that captures the major electoral dynamics that are relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work that also used factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.

@@

We limit our hypothetical elections to two candidates, to simplify the challenges of measuring harm with many candidates, over which voters may have different ranked preferences.
However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.

@@

We can also flip the setup and compute harm by averaging over the simulated plans and using the enacted plan as a counterfactual.
This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.

@@

This paper offers a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
The combination of these ideas imbues the analysis of districting plans with historical understandings of legal standing and harm.
This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.

>>>>

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank Steve Ansolabehere, Kosuke Imai, Gary King, Shiro Kuriwaki, 
      Tyler Simko, Nick Stephanopolous, and 
      participants of the Fall 2021 Harvard American Politics Research Workshop 
      and Imai Research Group for helpful comments and discussion.
      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: apsr
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].

Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
Many factors influence the performance and qualities of an electoral system.
Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
For example, in Massachusetts, it was impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
Once a set
of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

@@

A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022]. This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry]. The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}. Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.

@@

```{r minissouri, fig.width=6.75, fig.pos='b'}
#| fig.cap: |
#|      Two potential districting plans for Minissouri,
#|      a Democratic gerrymander (a), and a close-to-proportional plan (b).
#|      Panel (c) shows which voters are harmed by the Democratic gerrymander,
#|      using the proportional plan (b) as a counterfactual.
#|      Panel (d) summarizes this across precincts, showing
#|      the average harm done to voters of each partisan and racial group.
include_graphics("figures/minissouri.pdf")
```

@@

After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
These represent 44% of all Republicans and 2.5% of all Democrats.
We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
In direct terms, this means, on account of their partisan affiliation, Democrats would be 41.5% less likely to be represented by a Democrat than Republicans are to be represented by a Republican.

We need not limit our consideration to partisanship.
We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b), and it does so on the same scale as differential partisan harm.

@@

A voter is harmed only if they are not represented by the candidate of their choice under $q$ (i.e. $\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ (i.e. $\ind\{v(i) = \win_i(v, \tilde q)\}$).

Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$,
as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.

In general, $q$ harms some individuals while $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals.
A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.

@@

Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$. Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
In the algorithmic fairness literature, the implicit counterfactual is a perfect predictor and the actual decision process harms individuals whenever it makes a false positive or false negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
To quantify deviations from *fairness*, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
Appendix \ref{app:elecmodel} contains more details on how we estimate and apply this model.

More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
For example, one could operate on the logit scale, model incumbency effects, or add terms to account for local factors.

@@

With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
For each sampled counterfactual districting plan, $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.

@@

A primary advantage of our proposed harm measures is that they our measures are defined at the individual level. 
As such, we can study redistricting for any defined group memberships.
For social science, perhaps the two most important group memberships are that of race and party.
For these important contexts, we begin by showing that our measures provide an improvement over existing analytical approaches.

@@

The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
No simple formulas exist to measure the precise racial impacts of a districting plan.
Indeed, much of the academic literature has focused on the statistical challenges of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].

Modern analyses of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
Under the current framework for vote dilution established by @1986gingles, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.

While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, our proposed harm measures unifies all three *Gingles* criteria.
Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
But our proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.

@@

In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
We can account for geographic sorting, political cohesion, and the individual preferences of every voter.
Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.

@@

[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election. They are exactly the votes which don't contribute to winning a seat.

@@

Since none of the existing measures are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.

@@

Despite these differences, there are also empirical similarities in many cases.
The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.

@@

Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
In the worst cases, this can lead to directionally wrong conclusions, as we see in the next section.

@@

Figure \ref{fig:nj-pairs} in the appendix expands on Figure \ref{fig:nj-meas} to also show the correlations between these various measures.
Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.

@@

To apply our proposed framework to study the council district boundaries, we predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan, but city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R. Biden and Wu's opponent, Annissa Essaibi-George, on the right.
The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
Yet, the model is a useful simplification of reality which captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
The latent factor ecological inference model is not necessary for this kind of analysis; local survey data with ideological preferences and racial information would also suffice.

@@

Our hypothetical elections are focused on cases with two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.

@@

Recall that harm is asymmetric.
While that may be an undesirable property in some areas, here we can leverage that fact.
We flip the setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
Here, comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.

@@

This paper offers a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
The combination of these ideas imbues the analysis of districting plans with historical understandings of legal standing and harm.
This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.
In other words, your personal characteristics should not impact your ability to be represented.
